---
title: "Initial cross-system data comparison"
date: '2022-04-22'
output:
  pdf_document: default
  html_notebook: default
---

## General summary

- 849 fish were tagged across the four groups (coastal Massachusetts and the Hudson, Delaware, and Potomac Rivers) from 2014-2019
- 2016 was the only year where fish were concurrently tagged by each group
- Types of data collected were tag date, transmitter ID, external ID, in-house ID, total length, fork length, wet weight, scale age, sex, general area of tagging
- All tagging groups collected the date of tagging, age, and total length
  - Hudson fish are in the process of being aged
  - 15% Potomac fish have scales collected for ageing, but these have not yet been processed
- Hudson, Delaware, and Potomac River systems have sex data; coastal Massachusetts does not
- Hudson and Potomac have weights; Delaware and Massachusetts do not


```{r message=FALSE}
library(ggplot2); library(data.table)
```
```{r}
all_info <- tar_read(combined_tags)
all_info[, yr := as.factor(year(tagdate))]
all_info[, system := as.factor(system)]

all_info
```

## Systems

```{r}
addmargins(xtabs(~  system, addNA = T, data = all_info))
```

## Year tagged

- Potomac tags skew older, Delaware tags skew younger, Massachusetts tags are in the middle, and Hudson were only tagged in 2016

- 2016 is the only year that had fish tagged in all systems

```{r}
addmargins(xtabs(~ yr + system, addNA = T, data = all_info))
```

```{r}
ggplot(data = all_info) +
  geom_bar(aes(x = yr, fill = system))
```

## Age

- A subset of Potomac-tagged fish have not yet been, but are planning to be, aged. There will likely wind up being 1 Hudson River *NA* and 1 Potomac River *NA*.

```{r}
# Cross-tabulation
addmargins(xtabs(~ age + system, addNA = T, data = all_info))

# Range of ages
all_info[, .(min = min(age, na.rm = T),
             median = median(age, na.rm = T),
             max = max(age, na.rm = T)), 
         by = system]
```

- Delaware fish are oldest, followed by Hudson, then Potomac, then Massachusetts.

```{r}
TukeyHSD(aov(age ~ system, data = all_info))
```


```{r}
ggplot(data = all_info[!is.na(age)]) +
  geom_bar(aes(x = age, fill = yr)) +
  scale_fill_viridis_d() +
  facet_wrap(~ system) +
  labs(title = 'Age by year and system tagged')
```


```{r}
ggplot(data = all_info[!is.na(age)]) +
  geom_bar(aes(x = age, fill = sex)) +
  facet_wrap(~ system) +
  labs(title = 'Age by system tagged and sex')
```


## Sexes

Fish collected off the Massachusetts coast and 21% of Potomac fish were not sexed. "*NA*" means that sex was not recorded, while "*U*" means that the sex was not able to be determined.

```{r}
addmargins(xtabs(~ sex + system, addNA = T, data = all_info))
```

```{r}
ggplot(data = all_info) +
  geom_bar(aes(x = yr, fill = sex)) +
  facet_wrap(~ system)
```

## Lengths

All fish collected have total lengths, while only Hudson fish have fork lengths. Seems TL is the way to go here.

```{r}
ggplot(data = all_info) +
  geom_histogram(aes(x = tl, fill = yr), binwidth = 50) +
  scale_fill_viridis_d() +
  facet_wrap(~ system) +
  labs(title = 'Total length by year and tagged')
```

- Potomac and MA coast fish skew smaller, Hudson skews larger, and DE is bimodal (related in part to sex, see below)

```{r}
ggplot(data = all_info[!is.na(sex) & sex != 'U']) +
  geom_histogram(aes(x = tl, fill = yr), binwidth = 50) +
  scale_fill_viridis_d() +
  facet_grid(sex ~ system) +
  labs(title = 'Total length by sex and year and system tagged')
```


## Weights

Only Potomac and Hudson fish have recorded weights, with Potomac fish skewing much lighter.

```{r}
all_info[!is.na(wgt), .N, by = system]

ggplot(data = all_info[!is.na(wgt)]) +
  geom_histogram(aes(x = wgt, fill = yr), binwidth = 1000) +
  scale_fill_viridis_d() +
  facet_wrap(~ system)
```

